---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---

This is the first peer assessment for the Johns Hopkins Reproducible Research 
class on Coursera. Full details of the assignment can be found in the 
accompanying README.md file in this repository. Essentially, however, we load
some data captured from wearable tech and analyse this to answer the following
questions:

1. What is mean total number of steps taken per day?
2. What is the average daily activity pattern?
3. Are there differences in activity patterns between weekdays and weekends?

The following is a complete breakdown of the steps taken to perform this 
analysis, along with any relevant code, code output, plots and the results. 


## Loading and preprocessing the data

The first step in our analysis is to make sure we've got the raw data! 
The raw data was included in the repository when we forked this, but just in 
case, the code below will download/extract the raw data when and if necessary.

```{r}
# Download the activity.zip file if we don't already have this
dataUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zipFile <- "./activity.zip"
if (!file.exists(zipFile)) {
    download.file(dataUrl, zipFile, method="curl", mode = "wb")
}

# Extract the zip file if necessary
dataFile <- "./activity.csv"
if (!file.exists(dataFile)) {
    unzip(zipFile)
}
```

Then we load the activity data into a data frame tbl and assign this to the 
`activity` variable.

```{r, results='hide', message=F, warning=F}
library(dplyr)
library(ggplot2)
library(reshape2)
library(lattice)

activity <- tbl_df(read.csv(dataFile))
```

Now we're ready to perform some data analysis!


## What is mean total number of steps taken per day?

For this part of the anhalysis we ignore NA values. We start by creating a 
histogram of the total number of steps taken each day.

```{r}
# Create a daily summary of the activity data
daily <- activity %>%
    group_by(date) %>%
    summarise(steps=sum(steps, na.rm = T))

# Now plot a histogram showing the frequency with which various different step
# counts occur
hist(daily$steps, 
     main="Frequency distribution of daily steps", 
     xlab = "Number of Steps"
     )
```

From the histogram, we see that there is a clear right skew (which makes sense,
very few people are going to make a **lot** of steps each day).

Now we calculate the mean and median total number of steps taken per day.

```{r}
daily_mean <- mean(daily$steps)
daily_median <- median(daily$steps)
message(sprintf(
    "Mean daily steps: %.2f\nMedian daily steps: %d", daily_mean, daily_median
))
```


## What is the average daily activity pattern?

In this part of the assignment we try to see daily walking patterns so we look 
at the average number of steps taken during any part/interval of the day, over 
the course of the whole experiment.

We start by grouping the data appropriately. Then we make a time series plot 
(i.e. type = "l") of the 5-minute interval (x-axis) and the average number of 
steps taken, averaged across all days (y-axis).

```{r}
# Group the activity data by interval and summarise average daily activity 
# during that interval over the course of the experiment
by_interval <- activity %>%
    group_by(interval) %>%
    summarise(steps=mean(steps, na.rm = T))

# Make a quick line plot of the average daily activity
qplot(interval, steps, data=by_interval,
      geom="line",
      main = "Daily walking patterns (average steps ea. 5 minutes)"
      )
```

Then we calculate the **most active** part of the day, as defined by the 
5-minute interval, with the maximum average number of steps across all of the 
days in the experiment.

```{r}
peak_interval <- by_interval[which.max(by_interval$steps),]
message(sprintf(
    "Peak activity of %.2f steps during interval %d", 
    peak_interval$steps, 
    peak_interval$interval
    ))
```

## Imputing missing values

In this part of the assignment we identify periods with missing data and try to
impute values for those periods. 

We start by calculating the total number of missing values in the dataset (i.e. 
the total number of rows with NAs).

```{r}
partial_rows <- sum(!complete.cases(activity))
message(sprintf("%d rows are incomplete", partial_rows))
```

Now we implute values for those rows simply by taking the average value for the
corresponding 5-minute time interval and create a new dataset (`imputed) that is 
equal to the original activity dataset, only with the missing data filled in.

```{r}
imputed <- left_join(activity, by_interval, by="interval")
imputed[is.na(imputed$steps.x),1] <- imputed[is.na(imputed$steps.x),"steps.y"]
```

Now we make a histogram of the total number of steps taken each day and 
calculate the mean/median total number of steps taken per day using the imputed
data. 

```{r}
# Create a daily summary of the activity data
daily_imputed <- imputed %>%
    group_by(date) %>%
    summarise(steps=sum(steps.x))

# Now plot a histogram showing the frequency with which various different step
# counts occur
hist(daily_imputed$steps, 
     main="Frequency distribution of imputed daily steps", 
     xlab = "Number of Steps"
     )
imputed_daily_mean <- mean(daily_imputed$steps)
imputed_daily_median <- median(daily_imputed$steps)
message(sprintf(
    "Mean imputed daily steps: %.2f\nMedian imputed daily steps: %.2f", 
    imputed_daily_mean, 
    imputed_daily_median
))
```

Now we plot the imputed frequency distribution side by side with the original
frequency distribution to see the impact of imputing missing data on the 
estimates of the total daily number of steps.

```{r, message=F, warning=F}
# Collect the original and imputed data together in the same dataset
imputed_impact <- inner_join(daily, daily_imputed, by="date") %>%
    select(original=steps.x, imputed=steps.y)
# Now plot the two side by side
ggplot(
    melt(imputed_impact), 
    aes(value, fill = variable)
    ) + 
    xlab("steps") +
    ylab("frequency") +
    geom_histogram(position = "dodge")
```

As you might expect, by imputing values for missing values we substantially 
reduce the number of days in which we have very little data and (simply because
of the way we performed the imputation) we substantially increase the frequency 
of "mean" daily steps. The end result is that the imputed data looks a bit more 
normally distributed than the original data.

## Are there differences in activity patterns between weekdays and weekends?

For this part of the assignment we separate out activity on weekdays vs weekends
and then plot the results side by side, to guage the difference. *Note* that we 
use the imputed data for this analysis (i.e. there are no NA values).

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

We start by creating a new factor variable `day_type` in the dataset with two levels:
1. `weekday` and
2. `weekend`

Then we create two side by side plots for the subject's activity (average number 
of steps) in each 5-minute interval on both weekdays and weekends.


```{r}
imputed$day <- weekdays(as.POSIXct.Date(imputed$date))
imputed$day_type <- ifelse(imputed$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
work <- imputed %>% 
    group_by(day_type, interval) %>% 
    summarise(steps = sum(steps.x))
xyplot(steps ~ interval | day_type, 
       data = work, 
       layout = c(1, 2),
       xlab="Interval",
       ylab="Number of Steps",
       panel = function(...) {
            panel.xyplot(..., type = "l")
       }
       )

```

Clearly there's a bit more activity during weekdays early in the day (presumably 
going to work) and generally a lot more activity during the day on weekdays. The
subject is presumably a bit of a couch potato during the weekends!